\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}

\geometry{a4paper, margin=1in}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Multi-Tiered Candidate Matching Engine Architecture},
}

\title{\textbf{Multi-Tiered Candidate Matching Engine:\\ Architecture, Implementation, and Comparative Analysis}}
\author{PandyAi Engineering}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report details the architectural design and implementation of a three-tiered candidate matching engine designed to evaluate software engineering candidates against job descriptions. The system implements three distinct analytical methodologies: a rigid \textbf{Deterministic Baseline}, an embedding-driven \textbf{Hybrid Engine}, and an LLM-powered \textbf{Agentic Re-ranker}. This document provides a comprehensive technical breakdown of each approach, the underlying scoring mechanics, and a comparative analysis evaluating trade-offs between computational overhead, scalability, and semantic accuracy.
\end{abstract}

\tableofcontents
\newpage

\section{Introduction}
Modern Applicant Tracking Systems (ATS) struggle to accurately evaluate candidates due to the nuances in resume formatting, implicit skill transferability, and the rigidness of keyword-matching algorithms. To solve this, the developed matching engine transitions from a simple rule-based script into a modular, multi-tier orchestration pipeline.

By executing commands via the CLI (e.g., \texttt{--approach deterministic | hybrid | agentic}), the pipeline allows recruiters and system administrators to toggle between speed, semantic understanding, and deep contextual reasoning.

\section{System Architecture Overview}
The system is built upon a modular orchestration layer (\texttt{approaches.py}), standardizing the input/output boundaries for all evaluation strategies. Each job is processed alongside a pool of candidates, outputting two distinct artifacts per job:
\begin{enumerate}
    \item \textbf{Rank Data:} A sorted leaderboard of candidates with numerical scores.
    \item \textbf{Insights Data:} Natural language justifications, identified strengths, and actionable skill gaps.
\end{enumerate}
Global configurations (e.g., scoring limits, API keys, chunk sizes) are dynamically injected via \texttt{pydantic-settings} from an \texttt{.env} file.

\section{Approach 1: The Deterministic Baseline}
The Deterministic approach acts as the foundational layer. It utilizes strict string-matching algorithms, set intersections, and hard mathematical heuristics to calculate a score out of 100.

\subsection{Scoring Mechanics}
The scoring is divided into four main pillars:
\begin{itemize}
    \item \textbf{Must-Have Skills (50\%):} Pro-rated score based on the percentage of mandatory skills explicitly found in the candidate's reported skills. Missing a must-have triggers a severe algorithmic penalty multiplier (e.g., $0.25\times$ base score per missing skill) to heavily suppress underqualified candidates.
    \item \textbf{Nice-to-Have Skills (20\%):} Additive bonus points for supplementary technologies.
    \item \textbf{Experience Fit (25\% + 5\% Bonus):} Candidates meeting the required years of experience receive the baseline 25 points. Additional years yield bonus points (up to a capped maximum).
    \item \textbf{Location / Availability (5\%):} Points awarded based on logistical readiness and geographic proximity to the role.
\end{itemize}

\subsection{Limitations}
While highly performant ($\approx \mathcal{O}(N)$ computational complexity), the Deterministic engine fails to capture synonyms or implicit skills. For example, a candidate listing ``SCSS'' will be falsely penalized for lacking ``CSS''.

\section{Approach 2: The Hybrid Engine (Semantic Fusion)}
To address the keyword fragility of the Deterministic approach, the Hybrid engine introduces Natural Language Processing (NLP) via dense vector embeddings.

\subsection{Vectorization Pipeline}
Candidates and job descriptions are converted into high-dimensional vectors using the \texttt{BAAI/bge-small-en-v1.5} embedding model via the \texttt{sentence-transformers} library. The vector representations are stored locally in a \texttt{ChromaDB} persistent instance. The system maintains an in-memory ID cache hydrated from the disk to guarantee idempotent indexups and prevent redundant re-embeddings across multiple terminal sessions.

\subsection{Score Blending}
The Hybrid engine evaluates the candidate using the Deterministic logic, calculates the \textbf{Cosine Similarity} between the candidate's vector and the job's vector, and fuses the two scores mathematically:
\[ S_{hybrid} = (W_{det} \times S_{det}) + (W_{sem} \times S_{cos}) \]
Where $W_{det}$ (e.g., 0.7) and $W_{sem}$ (e.g., 0.3) are configurable weight parameters. This allows the system to algebraically rescue highly relevant candidates who suffer from poor keyword optimization.

\section{Approach 3: The Agentic Engine (LLM Re-Ranking)}
The pinnacle of the matching engine is the Agentic approach, which leverages Large Language Models (LLMs) to perform deep contextual reasoning, akin to a human recruiter.

\subsection{Phase 1: Hybrid Retrieval (Top-K)}
To prevent overwhelming the LLM's context window and to optimize API costs, the Agentic engine utilizes the Hybrid approach as a high-speed pre-retrieval mechanism. Only the top $K$ candidates (configurable via CLI flag) are passed to the next phase.

\subsection{Phase 2: LLM Evaluation and JSON Enforcement}
The top candidates are passed to an LLM (e.g., Llama-3 via the Groq API). A strict \texttt{system\_prompt} is injected, explicitly defining the 100-point limit boundaries (50 must-have, 25 experience, etc.) to ensure mathematical consistency with the Deterministic baseline. 
The LLM is forced to output a deeply nested JSON schema validated by \texttt{pydantic}. It dynamically infers skills (e.g., recognizing that building an authentication system implies knowledge of specific libraries), scores the candidates, and generates actionable "Next Steps" for candidate upskilling.

\subsection{Fault Tolerance}
Because massive batch processing can trigger LLM token limits or JSON truncation, the schema safely implements \texttt{Optional} fallback structures for deep insights. In the event of total API failure, the orchestration layer intercepts the error and elegantly falls back to the Hybrid engine outputs.

\section{Comparative Analysis}

\begin{table}[h]
\centering
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Metric} & \textbf{Deterministic} & \textbf{Hybrid} & \textbf{Agentic} \\ \midrule
\textbf{Processing Speed} & Instantaneous & Fast (CPU/GPU bound) & API Latency Dependent \\
\textbf{Computational Cost} & Zero & Low (Local Embeddings) & High (API Tokens) \\
\textbf{Semantic Awareness} & None & High (Similarity) & Maximum (Contextual) \\
\textbf{Explainability} & Transparent Math & Black-Box Vectors & Natural Language Justifications \\
\textbf{Implicit Skill Detection} & Fails & Moderate & Excellent \\ \bottomrule
\end{tabular}
\caption{Comparison of Candidate Evaluation Engines}
\end{table}

\subsection{Conclusion and Best Practices}
The multi-tiered architecture provides unparalleled flexibility. The \textbf{Deterministic} approach is ideal for rapid, strict pipeline filtering. The \textbf{Hybrid} approach offers the best balance of speed and semantic leniency, serving as the optimal default. Finally, the \textbf{Agentic} approach represents state-of-the-art matching accuracy, best reserved for final-stage shortlists where deep, human-like insight generation adds the highest value.

\end{document}
